Towards a standardized evaluation of imputation methodology

### 2022-03-31 Formatted/Formatted.tex
After removing comments, before fixing TODOs

% [Nowadays there are a lot of imp alg being propagated, esp ai/ml, whose perf often times seems quite promising. But may not be a fair comparison because of the simulation setup/dev in ai/ml so fast that steps may be skipped. Especially in the field of ML/AI, there are promising new imputation methods, which may yield even sharper imputations than now-standard (semi-)parametric imputation methods]

@@

After deciding upon a data-generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, it is not necessary to take sampling variation into account in order to obtain information about a method's ability to handle the missing data problem, or to objectively compare methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. A lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

% Table \ref{table:dgm_truth} displays the source of the comparative truth in simulation studies with and without sampling variance.When deviating conventional pooling rules \citep[cf.][p. 76-77]{rubi87} do not apply. Instead, alternative pooling rules need to be used \citep{raghunathan2003multiple,vink14}. If the sampling variance is not handled correctly, subsequent inferences may be invalid and the simulation results cannot be trusted.

@@

% Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions become increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions.

@@

% is the   [Other problems with missingness patterns: monotone/univariate vs multivariate, sporadic vs systematic, and univariately generated missing data may not have the expected multivariate patterns.] For example, iterative imputation algorithms converge instantly if there is only one variable with missingness or when the pattern is monotone, but require iteration in all other situations. And very specific missing data patterns may misguide which imputation method prevails in performance: some ad hoc methods are known to yield valid inferences under certain restricted conditions \citep[e.g., list-wise deletion may outperform imputation methods if there is missingness in the outcome variable of the analysis model exclusively][]{buur18}, while biasing inferences in other cases. 
% [Missingness mechanisms paragraphs: 1) definitions of mechanisms (now in table), 2) that MCAR is often ignored, while it's the minimum requirement and reference, and sometimes realistic in practice, 3) that MAR can be generated \textit{spuriously} if the correlation between variables is low,  4) that MNAR is often ignored, while it may be the most realistic in practice, and 5) finally add something about types of M(N)AR (ref Rianne's Dance paper).]
% MCAR missingness: The scenario where the missing values are missing at random and the observed values are observed at random. 
%  the validity of assumptions become increasingly important when the missingness increases.

@@

% Out of the three mechanisms , random missingness (MAR) is usually generated for the evaluation of imputation methodology. Missingness mechanisms that are not random (MNAR) and mechanisms that are considered non-ignorable \citep[see e.g.][]{rubi76}, are generally ignored, except for those methods that are specifically targeted at non-ignorable applications.

% MAR missingness is usually investigated in simulations and assumed in practice. Although MCAR might not reasonably be the true missingness generating mechanism and naturally not typically assumed, the mechanism  can be very insightful in the context of imputation methodology evaluation. By contrast, MNAR

% usually assumed to be random (MAR) or completely random (MCAR).

@@

% Figure \ref{fig:MAR} demonstrates two realizations of missingness mechanisms under MAR mechanisms derived from data with different multivariate relations.

% \begin{figure}[htb]
% \begin{center}
% \resizebox{\textwidth}{!}{
% \includegraphics[]{figures/spurious_MAR.pdf}}
% \caption{Three missingness mechanisms that yield approximately 50 percent missingness ($N=500$). Displayed are a MCAR mechanism, a right-tailed MAR mechanism generated from bivariate normal data with correlation $\rho=.8$ and the same MAR mechanism, now generated from bivariate normal data with correlation $\rho= 0$.}
% \end{center}
% \label{fig:MAR}
% \end{figure}
% \vspace*{1pc}

Finally, although one cannot definitively verify if the missingness is random--after all, for every MNAR mechanism there is a MAR mechanism with equal fit \citep{molenberghs2008every}--it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. As stated, this mechanism is not usually included in evaluations.  % The model--if any--that is used to generate the missingness is usually assumed to be random (MAR) or completely random (MCAR).

% Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions become increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions.

@@

% The evaluation criteria may depend on the estimand. When descriptive statistics are the goal and when statistical inference would not be of interest, bias of the estimates would still apply, but standard errors are generally ignored.

% Some simulators may focus on retrieving the unobserved values, while others study the performance in the context of obtaining efficient predictions or classification.  
% For many outcome-driven imputation tasks a single valid imputation may perhaps suffice.

% The evaluation criteria may depend on the estimand. When descriptive statistics are the goal and when statistical inference would not be of interest, bias of the estimates would still apply, but standard errors are generally ignored. Chances are that for those who focus on descriptive statistical applications, multiple imputation would not be the mode of choice.

% For example, a negligible absolute bias for a parameter for which the true value is zero, would yield infinite bias when relative bias is considered.

% [Add segway between evaluation of imputation and actual performance measures.]
% [Add that simulation aim/target determines which performance measures are relevant to consider, and that the choice of analysis model/estimand can influence the values of these metrics (e.g., focusing on univariate estimates only may favor ad hoc methods).]
% The choice of performance measures may inadvertently distort statistical properties of the imputed data. Developers often only inquire about the `accuracy' (i.e. how well can the method reproduce the original data). [Add RMSE at cell level here?]

@@

\begin{table}[tb]
\begin{center}
\caption{Steps to consider in imputation simulation studies.}
\label{table:steps}
\begin{tabular}{lll}
\hline
                        & Simulation aspects under consideration \\
\hline  
1. Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
2. Obtain truth         & Data-generating mechanism(s), estimand(s), sampling variance \\
3. Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
4. Apply methods        & Imputation method(s), analytic method(s) \\
5. Evaluate imputations & Algorithmic convergence, distributional characteristics, imputation \\
                        & model fit \\
6. Evaluate performance & Performance measures, qualitative judgment \\
7. Report               & Text, visualization, checklist, simulation script \\
\hline
\end{tabular}
\end{center}
% \footnotesize{Note. Multiple aspects may be varied across simulation conditions (e.g., using a factorial design with several data-generating mechanisms and missingness proportions).}
\end{table}

@@

% RMSE of the predicted values after imputation and estimation to say something about the comparative performance of the methods. Refer to missing data chapter. In this context, there is no talk of retrieving the `true' but unobserved values at all. The DGM is by definition design-based and there is no sampling variance. The missingness generation step is skipped and the only performance measure that can be used between methods is RMSE and maybe comparison with respect to CCA in terms of bias.]

@@

First, one should always consider MCAR missingness (see Table \ref{table:mech} for definitions). This mechanism should be used as a reference condition. Every imputation method should yield valid imputations under MCAR. We also know that, under MCAR, the distribution of imputed values should be equivalent to the distribution of observed values for a given variable, which may offer an indication of imputation model (mis)fit. [TODO: leg uit!]
% First, one should always consider MCAR missingness [as reference, because all imputation methods should yield valid inference under MCAR and we can use this mechanism to check imputation model fit] %, i.e. the scenario where the missing values are missing at random and the observed values are observed at random. % Under MCAR, the statistical properties of the observed data given the missing data are known and any imputation routine that cannot at least mimic the performance of the observed data inference, should be deemed inefficient in the scope of the simulation.

Next, missing data should be induced conform a model that is dependent on the observed data (i.e. a MAR mechanism). Arguably, MAR is the most often-assumed mechanism in practice, and should thus be evaluated carefully. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly} can be useful. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.% Given the simulated data distributions, one random missingness model may be far more disastrous to the observed information than another model. This may influence the performance of some (but not necessarily all) imputation routines. For example, inference from hot-deck techniques such as predictive mean matching \citep{little1988missing, rubin1986statistical} may be more severely impacted by large amounts of one-tailed missingness than inference from parametric techniques. It would be a shame to overlook results due to the focus on a single MAR mechanism.

@@

Apart from the missingness mechanisms, the missingness pattern must be varied. [TODO: beter bruggetje?] Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under 10\%, 25\% and 50\% univariate missingness (see Table \ref{table:prop}). % Although we limit the focus here to ignorable non-response, the suggested proportions are equally applicable to simulations under non-ignorable non-response.

@@

The imputation methods under evaluation should be applied to the generated incomplete data, after which an analysis model may be fitted to estimate the estimand(s). % It is important to choose the imputation model parameters carefully (e.g., number of imputations, number of iterations, regularization, etc.). After imputation, an analysis model may be fitted to estimate the estimand(s).
%[Use CCA as benchmark method: the imputation method should outperform CCA. Choose imputation model parameters carefully (number of imputations, number of iterations if applicable, predictor matrix, regularization if applicable, etc.). Estimate estimand(s) using the analysis model, use correct pooling rules.]
In addition, the simulator should always perform complete case analysis. We know the theoretical properties of complete case analysis, which makes the technique useful as a point of departure when evaluating imputation performance. Complete case analysis may, therefore, serve as a benchmark method against which imputation performance should be evaluated. Moreover, pairing complete case analysis with MCAR and MAR mechanisms allow the simulator to evaluate the validity of the missing data generation process under the chosen analysis model.% out-performing it should be a minimal requirement for the imputation method(s). %it is wise to evaluate the performance of complete case analysis (aka list-wise deletion) in all simulated conditions.

@@

Every imputation workflow should contain an evaluation of the obtained imputations. Even though inspecting each imputation may be labor-intensive due to the number of imputations generated in a simulation study, we highly recommend to consider the following aspects:
% A minimum requirement for any imputation method is the absence of non-convergence in the imputation-generating process. If non-convergence is suspected, the inference resulting from the imputations might be invalid. 
% [Some suggested minimal requirements for imputation methods: the absence of algorithmic non-convergence, the imputation model fits the observed parts of the data (PPC), under a MCAR mechanism the imputed values are distributed equivalent to the observed values.]

@@

We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values \citep[][\S 2.5]{buur18}. However, if a study is aimed at obtaining predictions or classification and inferential validity is not of interest, the RMSE of the predictions may yield valuable information about the methods' efficiency in terms of both accuracy and precision.
% [Some suggested minimum requirements for imputation methods: the method yields valid inference under MCAR, the method performs equally well or better than complete case analysis.]
% 
% % sim perf and model perf
% [The evaluation criteria may depend on the estimand. When descriptive statistics are the goal and when statistical inference would not be of interest, the bias of the estimates would still apply, but standard errors are generally ignored. Chances are that for those who focus on descriptive statistical applications, multiple imputation would not be the mode of choice.] 
% 
% 
% (i) Bias: Results should preferably be unbiased. However, the way bias is considered can greatly influence the interpretation of the results. For example, a negligible absolute bias for a parameter for which the true value is zero would yield infinite bias when relative bias is considered. The way bias is computed should therefore be carefully chosen and described. 
% (ii) Coverage: Coverage of a 95\% interval should in theory be $\geq 95$, where a coverage rate of 95\% would be most efficient. Under-coverage (when estimation is too liberal) may be an indication of invalid inference, while over-coverage (when estimation is too conservative) tells us that efficiency could still be gained. 
% (iii) Interval width: The way the confidence interval is calculated should be described. Wider intervals are associated with more uncertainty and the more narrow interval that is still properly covered indicates a sharper inference. However, inference from a wider interval that is properly covered is to be considered more valid when compared to a more narrow interval that is not properly covered anymore. 
% % [Add RMSE at cell level and at prediction level (if applicable).]

@@

% It could be convincing to demonstrate a method's applicability to real-world missing data problems. This can, for example, be achieved by obtaining and imputing a fully observed set of variables, wherein the missingness is mimicked from a similar, incomplete set. Alternatively, an incomplete set could be obtained and truth could be established by removing the incomplete cases from the data. However, the real-world missingness would then be omitted. [TODO: check waar dit al eerder staat!]

@@

% [Be explicit in choices. Describe the simulation textually, use pseudo-code, a flowchart, visualize missingness and simulation results. For example, the procedure to generate missing data is often insufficiently described in publications. The used missingness mechanism should be detailed, either graphically or written as a function of the data. Fill in the following checklist (in appendix).]

@@

% \subsection*{A.1.\enspace Simulation setup}

% Let $Y=(Y_{obs},Y_{mis})$ be an incomplete semicontinuous variable, where $Y_{obs}$ and $Y_{mis}$ denote the observed values and the missing values in $Y$, respectively. Further, $X=(X_1,\dots,X_k)$ is a set of $k$ fully observed covariates, where $X_{obs}$ and $X_{mis}$ correspond to the observed an missing parts in $Y$ and where $X_j$ would indicate the $j$th variable in $X$, with $j=1,\dots, k$. Finally, let $R$ be a response indicator that is 1 if $Y$ is observed and 0 if $Y$ is missing.
% 
% To prove the point, we generated a single data set from the multivariate normal distribution with means
% \begin{equation*}\label{meanmatrix}
% \mu= \bordermatrix{&	 \cr
%  Y	&5	\cr
% { X}_1	&10	\cr
% { X}_2	&10	},
% \end{equation*}
% and covariance matrix
% \begin{equation*}\label{covmatrix}
% \Sigma = \bordermatrix{& Y &{ X}_1 &{ X}_2	 \cr
%  Y	&1	&0.8	&0\cr
% { X}_1	&0.8	&1	&0\cr
% { X}_2	&0	&0	&1}.
% \end{equation*}
% 
% Three incomplete sets were then generated, wherein random missingness was imposed in $Y$ in three ways. First, we left the observed data \emph{observed at random} to impose MCAR missingness following the mechanism 
% \begin{equation*}
% P(R=0|Y_{obs},Y_{mis}, X_j)=P(R=0)=.50.
% \end{equation*}
% Next, I created random missingness in $Y$ according to the following MAR missingness mechanism
% \begin{equation*}
% P(R=0|Y_{obs},Y_{mis}, X_j)=P(R=0|Y_{obs}, X_j),
% \end{equation*}
% by using a random draw from a binomial distribution of the same length as $Y$ and of size 1 with missingness probability equal to the inverse logit
% \begin{equation*}
% P(R=0)=\frac{e^{a}}{(1+e^{a})}.
% \end{equation*}
% where $a=(\mu_{X_j}-X_{ij})/\sqrt{\mathbb{E}[X_{ij} - \mu_{X_j}]}$ in order to create approximately 50\% right-tailed MAR missingness. I followed this procedure twice; once for each $X_j$.

% The three resulting incomplete data sets were then imputed with package \texttt{mice} \citep[][v2.25]{mice} in \texttt{R} \citep[][v3.2.3]{R}, with Bayesian linear regression imputation (\texttt{mice.impute.norm}) as the imputation routine. The completed-data analyses were combined into a single inference following the rules described in \citet{vink2014pooling}.

>>>>

@@

After deciding upon a data-generating mechanism, another influential--but often overlooked--decision awaits. Simulation studies on missing data methodology have the unique option to exclude sampling variance from their evaluations and only use the missing data generation procedure as the source of Monte Carlo variation. After all, we are interested in the missingness and are not considering the noise induced by the sampling mechanism for evaluation in such studies. Therefore, taking sampling variation into account is neither necessary for obtaining information about a method's ability to handle the missing data problem, nor for objectively comparing methods on their ability to correct for missingness \citep[see for a detailed discussion][]{vink14}. The only required change to simulation setups is to remove sampling variance from performance evaluation. Still, a lot of published work does not consider this option, while it could have sharpened inconclusive or obfuscated results.

@@

@@

@@

@@

Finally, although one cannot definitively verify if the missingness is random--after all, for every MNAR mechanism there is a MAR mechanism with equal fit \citep{molenberghs2008every}--it can be argued that MNAR is the more likely mechanism for real-life missingness scenarios. As stated, this mechanism is not usually included in evaluations.

@@

@@

\begin{table}[tb]
\begin{center}
\caption{Steps to consider in imputation simulation studies.}
\label{table:steps}
\begin{tabular}{lll}
\hline
                        & Simulation aspects under consideration \\
\hline  
1. Set scope            & Aim(s), missing data method(s) to evaluate, simulation design \\
2. Obtain truth         & Data-generating mechanism(s), estimand(s), sampling variance \\
3. Induce missingness   & Missingness mechanism(s), missing data pattern(s) \\
4. Apply methods        & Imputation method(s), analytic method(s) \\
5. Evaluate imputations & Algorithmic convergence, distributional characteristics, imputation \\
                        & model fit \\
6. Evaluate performance & Performance measures, qualitative judgment \\
7. Report               & Text, visualization, checklist, simulation script \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

@@

First, one should always consider MCAR missingness (see Table \ref{table:mech} for definitions). This mechanism should be used as a reference condition. Every imputation method should yield valid imputations under MCAR. We also know that, under MCAR, the distribution of imputed values should be equivalent to the distribution of observed values for a given variable, which may offer an indication of imputation model (mis)fit. [TODO: leg uit!]

Next, missing data should be induced conform a model that is dependent on the observed data (i.e. a MAR mechanism). Arguably, MAR is the most often-assumed mechanism in practice, and should thus be evaluated carefully. A straightforward technique for inducing univariate MAR missingness is described in \citet[][\S 3.2.4]{buur18}, generalizations to multivariate MAR missingness can be found in \citet{ampute}. If the missingness is to be induced in longitudinal data autoregressive MAR models can be useful \citep[see e.g.][model 2 and model 3]{shara2015randomly} can be useful. It is advisable to investigate varying shapes of MAR missingness to achieve a more realistic indication of the robustness of the imputation performance across the range of random missingness. The effects of different types of MAR mechanisms are described in \citet{scho18}.

@@

Apart from the missingness mechanisms, the missingness pattern must be varied. [TODO: beter bruggetje?] Remember that missingness is only ignorable under MAR when the parameter of the data is distinct and a-priori independent from the parameter of the missing data process. Under MAR missingness we assume that we may use the observed data to make inferences about the joint (observed and unobserved) data. The dependency of the procedure on the assumption under which we obtain inference is only influenced by the amount of missingness. If there is no missingness--or if there is no data, for that matter--the inference does not depend on the assumption. Alternatively, the validity of assumptions becomes increasingly important when the missingness increases. Since we control the MAR mechanism, the assumption under which we may solve the missing data problem should hold and it is only fair to assess performance under stringent missingness conditions. We, therefore, propose to evaluate imputation methodology under 10\%, 25\% and 50\% univariate missingness (see Table \ref{table:prop}).

@@

The imputation methods under evaluation should be applied to the generated incomplete data, after which an analysis model may be fitted to estimate the estimand(s). In addition, the simulator should always perform complete case analysis. We know the theoretical properties of complete case analysis, which makes the technique useful as a point of departure when evaluating imputation performance. Complete case analysis may, therefore, serve as a benchmark method against which imputation performance should be evaluated. Moreover, pairing complete case analysis with MCAR and MAR mechanisms allow the simulator to evaluate the validity of the missing data generation process under the chosen analysis model.

@@

Every imputation workflow should contain an evaluation of the obtained imputations. Even though inspecting each imputation may be labor-intensive due to the number of imputations generated in a simulation study, we highly recommend to consider the following aspects:

@@

We do not generally recommend the use of the RMSE as evaluation criterion, because this metric does not account for the inherent uncertainty of the missing values \citep[][\S 2.5]{buur18}. However, if a study is aimed at obtaining predictions or classification and inferential validity is not of interest, the RMSE of the predictions may yield valuable information about the methods' efficiency in terms of both accuracy and precision.

@@

% It could be convincing to demonstrate a method's applicability to real-world missing data problems. This can, for example, be achieved by obtaining and imputing a fully observed set of variables, wherein the missingness is mimicked from a similar, incomplete set. Alternatively, an incomplete set could be obtained and truth could be established by removing the incomplete cases from the data. However, the real-world missingness would then be omitted.

@@

@@
